tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-08 paper.tex
First pass at full intro

% ARG is especially useful in population genetic simulations, and it would be
% possible to simulate genetic data with ancestral information by using ARG.
% Simulation is used in GWAS to examine how population structure and demographic
% histories influence polygenic risk score prediction
% \citep{martin2017,zaidi2020}, because simulation is the only way to obtain
% genetic data where the underlying demographic information is known. A succinct
% tree sequence (or tree sequence) is a data structure for ARGs, and it is one of
% the most popular genomic data structures in population genetic simulations
% \citep{baumdicker2022,haller2019}, (cite "A general and efficient, ..."
% article). Since it can efficiently store genetic information with far less data
% size than other methods, it is an appealing choice for simulating mutation and
% ancestry information \citep{kelleher2019}.

There are multiple useful software packages to simulate genotypic data with
quantitative traits \citep{gaynor2021,haller2023}, and to simulate quantitative
traits from observed genetic sequences \citep{fernandes2020,meyer2018}.
However, there does not exist any simulators that can efficiently simulate
quantitative traits of individuals directly from an ARG. While it would be
possible to convert a tree sequence data to a different data format and use the
available packages to simulate quantitative traits, data conversion can be
computationally costly and it diminishes the computational advantages of using
tree sequence data. It would be much more efficient to analyze genetic
information by using the tree sequence data structure, and many population
genetic statistics can be efficiently computed from the tree structure instead
of obtaining the genotype matrix \citep{ralph2020}.

In this paper, we propose a new Python library \texttt{tstrait} to simulate
quantitative traits of individuals in an arbitrary ARG. \texttt{tstrait} is
built on top of \texttt{tskit} \citep{ralph2020}, which provides an efficient
way for analyzing and processing ARGs encoded as tree sequence.
\texttt{tstrait} utilizes the efficient data structure in the tree sequence
data to simulate complex traits at a fast computational speed. It also provides
an easily accessible interface by using \texttt{pandas} dataframe
\citep{pandas} as an input and output of the simulation. Thus, \texttt{tstrait}
can be easily integrated with other population genetic simulators as well.

In addition to its computational advantages of processing tree sequence data,
\texttt{tstrait} offers other advantages: \emph{i)} it can simulate traits even
if the sites have multiple mutations, unlike other simulation frameworks that
assume infinite-sites model \citep{martin2017}, \emph{ii)} it can simulate
traits from any ancestral individuals that existed in the demographic history,
unlike other packages that can only simulate traits from the observed genetic
information \citep{fernandes2020,meyer2018}, and \emph{iii)} it integrates with
Python data science packages, such as \texttt{NumPy} \citep{numpy} and
\texttt{pandas} \citep{pandas}. Due to these advantages, we hope that
\texttt{tstrait} can provide easily accessible state-of-art quantitative trait
simulation for researchers who are using population genetic simulation tools.

>>>>

An important property of the succinct tree sequence ARG encoding
used by these population genetic simulators is that it allows us to
store vast genetic variation datasets very compactly. Rather than
storing the genetic sequence variation in its conventional matrix
form, we can instead store the underlying genealogical history
and mutations that ultimately \emph{result in} the observed
variation~\citep{kelleher2019}.
For example, the ultra-realistic simulations of 1.4 million
French-Canadian individuals provided by \citet{anderson2023}
requires around 550 GB of storage space in the
standard gzip-compressed VCF format ($\sim$280TB uncompressed)
for chromosome 9. The original simulation file representing
precisely the same genetic variation data, \emph{plus} the
complete simulated ARG, requires only 1.36GB of space
in the compressed \texttt{tszip} format. Thus, in this
case we achieve more than 400-fold compression,
and this massive and otherwise highly unwieldy dataset
becomes straightforward to work with (we discuss
landmark dataset again in later sections).
Another substantial benefit of storing simulated datasets
in this ARG-based format is that many computations can
be expressed efficiently in terms of the underlying
ancestry~\citep{kelleher2016efficient,ralph2020}, without
needing to decode the actual variation data.

The first step of simulating data for evaluating GWAS methods,
generating realistic genetic variation datasets of millions
of individuals, is therefore at a high-level of sophistication
and more than sufficient to support ongoing research.
The second step, of generating the actual quantiative traits,
is currently lagging far behind.
Although there are multiple packages to simulate genotypic data
concurrently with quantitative traits~\citep{gaynor2021,haller2023},
and to simulate quantitative traits from a given set of genetic
sequences~\citep[e.g.][]{meyer2018,fernandes2020}, none can
handle the millions of samples now required.
In particular, there is no software package available to simulate quantitative
traits directly from the efficient ARG representation used by
modern population genetic simulators.

In this paper we present \texttt{tstrait}, a new Python library that
efficiently simulates quantitative traits on an arbitrary ARG.
The package is build on top of
\texttt{tskit}~\citep{ralph2020}, which provides a powerful
framework for storing and processing ARGs.
Using these efficient data structures, \texttt{tstrait} can
quickly simulate quantitative traits for population-scale datasets,
with a very low memory overhead. This approach also provides a
richer model than existing methods,
allowing us to, for example, simulate traits for
ancestral individuals as well as contemporary samples.
The \texttt{tstrait} library also integrates well with the wider
PyData ecosystem of data-science tools,
using
\texttt{NumPy} arrays~\citep{numpy} and
\texttt{pandas} dataframes~\citep{pandas}
as input and output.
